Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-12 manuscript/manuscript.tex
Improve D determination text

\begin{table}
\caption{
    Description of the tesseroid collections used to build a spherical shell and
    characterize the accuracy of the numerical integration.
    The first three columns describe the shapes of the tesseroid collections: each one
    specifies the number of tesseroids on the corresponding dimension (radial,
    latitudinal, longitudinal).
    The last two columns specify the latitudinal and longitudinal size of each
    tesseroid that compose the spherical shell model.
    For each collection, we define four spherical shell models with an outer radius
    equal to the mean Earth radius (6378.137 km) and the inner radius determined by the
    following thicknesses: 100m, 1km, 10km, 100km and 1000km.
}
\label{tab:tess-models}
\begin{tabular}{cccccc}
    Radial & Latitudinal & Longitudinal  & Total tesseroids & $L_\phi$ & $L_\lambda$
    \\ \hline
    1 & 1  & 2  & 2   & 180$^\circ$ & 180$^\circ$ \\
    1 & 6  & 12 & 72  &  30$^\circ$ &  30$^\circ$ \\
    1 & 12 & 24 & 288 &  15$^\circ$ &  15$^\circ$ \\
    1 & 18 & 36 & 648 &  10$^\circ$ &  10$^\circ$ \\
\end{tabular}
\end{table}

@@

In order to compare the numerical results with the analytical solution we
must build spherical shell models made of tesseroids.
These shell models are determined by their inner and outer radii and by the collection
of tesseroids used to build the shell.
For this analysis we have chosen to set the outer radii of every shell model equal to
the mean Earth radius (6378.137 km), while the inner radius is determined by the desired
thickness of the shell.
We choose  different values of shell thickness (100m, 1km, 10km, 100km and 1000km) in
order to account for different computation scenarios.
Furthermore, we propose different tesseroids collections to build the shell model by
regularly splitting it along the longitudinal and latitudinal directions.
Each collection, described in Table~\ref{tab:tess-models}, is composed by a different
number of tesseroids, i.e. with tesseroids of different sizes.
Finally, for each collection we build a shell model with the thicknesses
previously described, giving a total of 20 different shell models.

@@

We assign a density that varies linearly with $r$ (Eq.~\ref{eq:density-linear}) to each
shell, which assumes the values of $\rho_\text{in} = 3300$~kg/m$^3$ and
$\rho_\text{out} = 2670$~kg/m$^3$ on the inner and outer radii, respectively.
So, the constants $a$ and $b$ from Eq.~\ref{eq:density-linear} must be

@@

\noindent
where $R_\text{out}$ and $R_text{in}$ are the outer and inner radii of the shell,
respectively.
For every shell defined in Table~\ref{tab:tess-models}, the outer radius will be equal
to the mean Earth radius (6378.137 km).

@@

We compute the gravitational potential ($V$) and the vertical component of the gradient
($g_z$) that each shell model detailed on Table~\ref{tab:tess-models} generates on each
grid defined in Table~\ref{tab:grids}.
Other components of the gradient are equal to zero outside of the shell due to the
rotational symmetry and are thus omitted from the analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 5 with a step of
0.5.
The differences are shown in Fig. \ref{fig:D-linear} as relative to the spherical shell
values.
We omit the differences for $g_{xx}$ and $g_{yy}$ because they are equivalent to the
ones obtained for $g_{zz}$.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

We observe from Fig. \ref{fig:D-linear} that the relative errors for the
potential and the $g_z$ component fall below the 0.1\% threshold at
$D=1$ and $D=2$, respectively, for both shell thicknesses.
On the other hand, $g_{zz}$ displays a noticeable
difference between the thin and the thick shell: the first one needs a
value of $D$ equal to 8 while the later only a $D$ of 3.

>>>>

\begin{table}
\caption{
    Description of the tesseroid collections used to build a spherical shell and
    characterize the accuracy of the numerical integration.
    The first three columns describe the shapes of the tesseroid collections: each one
    specifies the number of tesseroids on the corresponding dimension (radial,
    latitudinal, longitudinal).
    The last two columns specify the latitudinal and longitudinal size of each
    tesseroid that compose the spherical shell model.
    For each collection, we define four spherical shell models with an outer radius
    equal to the mean Earth radius (6378.137 km) and the inner radius determined by the
    following thicknesses: 100m, 1km, 10km, 100km and 1000km.
}
\label{tab:tess-collections}
\begin{tabular}{cccccc}
    Radial & Latitudinal & Longitudinal  & Total tesseroids & $L_\phi$ & $L_\lambda$
    \\ \hline
    1 & 1  & 2  & 2   & 180$^\circ$ & 180$^\circ$ \\
    1 & 6  & 12 & 72  &  30$^\circ$ &  30$^\circ$ \\
    1 & 12 & 24 & 288 &  15$^\circ$ &  15$^\circ$ \\
    1 & 18 & 36 & 648 &  10$^\circ$ &  10$^\circ$ \\
\end{tabular}
\end{table}

@@

In order to compare the numerical results with the analytical solution we
must build spherical shell models made of tesseroids.
These shell models are determined by their inner and outer radii and by the collection
of tesseroids used to build the shell.
For this analysis we have chosen to set the outer radii of every shell model equal to
the mean Earth radius (6378.137 km), while the inner radius is determined by the desired
thickness of the shell.
We choose  different values of shell thickness (100m, 1km, 10km, 100km and 1000km) in
order to account for different computation scenarios.
Furthermore, we propose different tesseroids collections to build the shell model by
regularly splitting it along the longitudinal and latitudinal directions.
Each collection, described in Table~\ref{tab:tess-collections}, is composed by
a different number of tesseroids, i.e. with tesseroids of different sizes.
Finally, for each collection we build a shell model with the thicknesses
previously described, giving a total of 20 different shell models.

@@

We assign a density that varies linearly with $r$ (Eq.~\ref{eq:density-linear}) to each
shell model, which assumes the values of $\rho_\text{in} = 3300$~kg/m$^3$ and
$\rho_\text{out} = 2670$~kg/m$^3$ on the inner and outer radii, respectively.
So, the constants $a$ and $b$ from Eq.~\ref{eq:density-linear} must be

@@

\noindent
where $R_\text{out}$ and $R_\text{in}$ are the outer and inner radii of the shell,
respectively.
Because the outer radius of every spherical shell model will be equal to the mean Earth
radius, $R_\text{out} = 6378.137$km.

@@

We compute the gravitational potential ($V$) and the vertical component of the gradient
($g_z$) that each shell model generates on each grid defined in Table~\ref{tab:grids}.
Other components of the gradient are equal to zero outside of the shell due to the
rotational symmetry and are thus omitted from the analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 5 with a step of
0.5.
Then we calculate the absolute difference between these results and the analytical
solution, taking only the maximum absolute difference on each computation grid.
This differences are shown in Fig.~\ref{fig:D-linear}.
In order to simplify the visualization of the results, only the maximum
differences corresponding to tesseroids collections with equal thickness are shown
in the Figure, obtaining one curve for every shell thickness.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

We observe from Fig. \ref{fig:D-linear} that the relative errors for the
potential and the $g_z$ component fall below the 0.1\% threshold at
$D=1$ and $D=2.5$, respectively, for every shell model and computation grid.
These results also show that there's no meaningful dependency between the accuracy of
the method and the thickness of the shell, the collection of tesseroids used to build
it or the height of the computation grid (the results for the Satellite grid and the
Global grid are very similar).
